hatchR: A toolset to predict hatch and emergence phenology in wild fishes

### 2025-03-03 inst/manuscript/paper.Rmd
Case Study 1 edits

# Case Study 1:

**Bull Trout in the Crooked River** OR **Predicting Emergence Timing for Management Actions**

@@

In this scenario, we consider road work near the upper portion of Crooked River in the Boise River watershed, where a key Bull Trout (*Salvelinus confluentus*) population exists. Bull Trout, a federally threatened species under the Endangered Species Act [@nolfi2024], are particularly sensative to sediment disturbance. The Forest Service Fisheries Biologist overseeing the project wants to determine whether Bull Trout fry will likely be out of the gravel and free-swimming by June 1st. In this system, Bull Trout typically complete spawning by the end of September, so we consider the latest possible spawn date: September 30th.

@@

Once uploaded, **hatchR** automatically generates a vizual data check using `plot_check_temp()`. After confirming data integrity, users navigate to the `Model Phenology` window. For this case study, we use the preloaded Bull Trout parameterization from @austin2019, selecting the `Existing` model option via dropdown menus. The user then chooses multiple spawn dates using an interactive calendar. Here, we focus on September 30th as outlined in our management scenario.

@@

In this example, the model predicts that the last Bull Trout will emerge before the June 1st traget date. This suggests that the Fisheries Biologist can confidently approve the road work in the area without concern for sediment disturbance impacting fish developing in the gravel. This type of predictive modeling helps managers make informed, science-based decisions that balance conservation priorities with necessary land-use activities.

>>>>

# Case Study 1

**Predicting Emergence Timing for Management Actions**

@@

In this scenario, we consider road work near the upper portion of Crooked River in the Boise River watershed, home to a key Bull Trout (*Salvelinus confluentus*) population. Bull Trout, a federally threatened species under the Endangered Species Act [@nolfi2024], are particularly sensative to sediment disturbance. The Forest Service Fisheries Biologist overseeing the project wants to determine whether Bull Trout fry will likely be out of the gravel and free-swimming by June 1st. In this system, Bull Trout typically complete spawning by the end of September, so we consider the latest possible spawn date: September 30th.

@@

Once uploaded, **hatchR** automatically generates a vizual data check using `plot_check_temp()`. After confirming data integrity, users navigate to the `Model Phenology` window. For this case study, we use the pre-loaded Bull Trout parameterization from @austin2019, selecting the `Existing` model option via dropdown menus. The user then chooses multiple spawn dates using an interactive calendar. Here, we focus on September 30th (in the 2014 spawn year) as outlined in our management scenario.

@@

In this example, the model predicts that the last Bull Trout will emerge before the June 1st road work target date. This suggests that the Fisheries Biologist can confidently approve the work in the area without concern for sediment disturbance impacting fish developing in the gravel. This type of predictive modeling helps managers make informed, science-based decisions that balance conservation priorities with necessary land-use activities.
